\LARGE \bf Analytic Test Functions for Generalizable \\ Evaluation of Convex Optimization Techniques

### 2020-03-30 [2019-11]_HPC_(quintic_spline)/paper/paper.tex
updated work

@@

An algorithm for computing monotone order six piecewise polynomial
interpolants is proposed. Algebraic constraints for enforcing
monotonicity are provided that align with quintic monotonicity
theory. The algorithm is implemented, tested, and applied to several
sample problems to demonstrate the improved accuracy of monotone
quintic spline interpolants compared to that for existing monotone
cubic spline interpolants.

@@

Many domains of science rely on smooth approximations to real-valued functions over a closed interval. Piecewise polynomial functions (splines) provide the smooth approximations for animation in graphics \cite{herman2015techniques,quint2003scalable}, aesthetic structural support in architecture \cite{brennan2019measure}, efficient aerodynamic surfaces in automotive and aerospace engineering \cite{brennan2019measure}, prolonged effective operation of electric motors \shortcite{berglund2009planning}, and accurate nonparametric approximations in statistics \cite{knott2012interpolating}. While polynomial interpolants or regressors apply broadly, splines are often a good choice because they can approximate globally complex functions while minimizing the local complexity of an approximation.

@@

The importance of piecewise quintic interpolation over lower order approximations can be simply demonstrated. In general, the order of a polynomial determines the number of function values it can interpolate, and the growth rate of error away from the interpolated function values. As demonstrated in Figure \ref{fig:spline_demonstration}, it can be seen that matching a value at either end of the interval requires an order two (linear) approximation and each additional derivative at the ends of the interval raises the necessary polynomial order by two. The body of this work is composed of a novel algorithm for enforcing monotonicity on quintic polynomial pieces, then extending that solution to work on quintic splines.

@@

The current state-of-the-art monotone interpolating spline with a mathematical software implementation is piecewise cubic, continuously differentiable, and was first proposed in \cite{fritsch1980monotone} then expanded upon in \cite{carlson1985monotone}. Let $\pi: x_0 = k_1 < k_2 < \cdots < k_n = x_1$ be a partition of the interval $[x_0,x_1]$. Let $f: \mathbb{R} \rightarrow \mathbb{R}$ be $C^2$, and $\bigl\{f(k_i)\bigr\}_{i=1}^n$ and $\bigl\{\Delta_i\bigr\}_{i=1}^n$ be given sets of function and derivative values at the partition points for a monotone function $f$. Either $f(k_i) \leq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\ge0$, $i=1$, $\ldots$, $n$, or $f(k_i) \geq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\le0$, $i=1$, $\ldots$, $n$. Let $\hat f$ be a piecewise cubic polynomial defined in each subinterval $I_i = [k_i, k_{i+1}]$ by

\begin{align*}
h_i &= k_{i+1} - k_{i}, \quad
u(t) = 3t^2 - 2t^3, \quad
p(t) = t^3 - t^2,\\
\hat f(x) &= f(k_i) u\big((k_{i+1} - x) / h_i\big) +
f(k_{i+1}) u\big((x - k_i) / h_i\big) \\
&{} - h_i\Delta_i p\big((k_{i+1}-x)/h_i\big) + h_i\Delta_{i+1}
p\big((x-k_i)/h_i\big).
\end{align*}

Notice that a trivially monotone spline results when $\Delta_i = 0$, for $i = 1$, $\ldots$, $n$. However, such a spline has too many \textit{wiggles} for most applications. \citeN{carlson1985monotone} show that simple conditions on the derivative values can guarantee monotonicity, and that these conditions can be enforced in a way that ensures modifications on one interval will not break the monotonicity of cubic polynomials over any neighboring intervals. Consider the terms $\alpha = \big(\Delta_i (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$ and $\beta = \big(\Delta_{i+1}(k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$. If $f(k_i) = f(k_{i+1})$, take $\Delta_i = \Delta_{i+1} =0$ and $\alpha =\beta =1$. Monotonicity of a cubic polynomial over a subinterval can be maintained by ensuring that $\alpha$ and $\beta$ reside in any of the regions depicted in Figure \ref{fig:projection}.

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{cubic_projection_demonstration}
  \caption{These are the feasible regions of monotonicity for cubic splines and the projections that make a cubic polynomial piece monotone. The regions themselves are numbered 1--4 corresponding to their original description in Fritsch and Carlson (1980). One point is projected onto each region as a demonstration.}\label{fig:projection}
\end{figure}

@@

The following section is composed of three algorithms that together are used to construct a monotone quintic spline interpolant. Without loss of generality, the algorithms will only consider the monotone increasing (nondecreasing) case. The monotone decreasing (nonincreasing) case is handled similarly. Algorithm \ref{alg:check-monotone} checks monotonicity,
Algorithm \ref{alg:make-monotone} enforces monotonicity on an order six
polynomial piece of the quintic spline, and Algorithm
\ref{alg:monotone-spline} uses the previous two algorithms to enforce
monotonicity for the entire quintic spline.

@@

These algorithms make use of first and second derivatives of the approximated function. In the case that the first and second derivative information is not provided along with function values, the derivatives are estimated with finite differences of the function values. The final quintic spline is represented as a piecewise polynomial using the Newton form for each polynomial piece, and ultimately converted to a $B$-spline representation for evaluation.

@@

Notice that both Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} have $\mathcal{O}(1)$ runtime  assuming a fixed level of relative precision. A constant number of operations are needed for verifying monotonicity (Algorithm \ref{alg:check-monotone}), while a constant set of operations and a single binary search are performed for enforcing monotonicity (Algorithm \ref{alg:make-monotone}). The search requires a fixed number of steps to achieve any predetermined relative precision. Also note that an efficient implementation of Algorithm \ref{alg:make-monotone} only needs to recalculate Steps \ref{alg1-abg} and \ref{alg1-last} of Algorithm \ref{alg:check-monotone} during the binary search. Next, an algorithm for constructing a monotone quintic spline interpolant is presented.

@@

\algorithm
    {monotone\_spline$\big((k_1$, $\ldots$, $k_n)$, $f$, $s\big)$}
    {where $(k_1$, $\ldots$, $k_n)$ is an increasing sequence of real numbers, $f$ is an order six piecewise polynomial with breakpoints $k_1$, $\ldots$, $k_n$ defined by the data $\bigl\{f(k_i)\bigr\}_{i=1}^n$, $\bigl\{Df(k_i)\bigr\}_{i=1}^n$, $\bigl\{D^2f(k_i)\bigr\}_{i=1}^n$, and $s > 1$ is an integer shrink factor.}
    {\label{alg:monotone-spline}
    \item create queue
      \\ create changed$(1$, $\ldots$, $n) := $ FALSE
      \\ create step\_size$(1$, $\ldots$, $n) := \bigl(Df(k_1)/s$, $\ldots$, $Df(k_n)/s\bigr)$ \label{step:create}
      \comment{The three variables defined in Step \ref{step:create} are used to ensure the eventual achievement of monotonicity. The \texttt{queue} is a standard first in first out (FIFO) queue with \texttt{enqueue} and \texttt{dequeue} operations. The array \texttt{changed} contains Booleans describing whether or not a breakpoint belongs to an interval that has been modified to enforce monotonicity. The \texttt{step\_size} array contains the real-valued derivative decrement step sizes to use in the search for a valid monotone spline.}
    \item for $i := 1$, $\ldots$, $n-1$
      \indented{enqueue $\bigl((k_i, k_{i+1})\bigr)$}
      \comment{Initially, all intervals must be checked for monotonicity. The following loop will continue until all intervals are verified as monotone without need for modification.}
    \item while $($size$($queue$) > 0)$
    \item \indented{$(x_j$, $x_{j+1}) := $ dequeue}
    \item \indented{if $\big($not is\_monotone$(x_j$, $x_{j+1}$, $f)\big)$}
    \item \indented{\indented{if $\big($changed$(j)\big) \qquad \ Df(x_j) := Df(x_j) -{}$step\_size$(j)$}}
      \indented{\indented{if $\big($changed$(j+1)\big) \quad Df(x_{j+1}) := Df(x_{j+1}) -{}$step\_size$(j+1)$}}
      \indented{\indented{\comment{If a breakpoint belongs to an interval that has been previously modified, then the enforced monotonicity conditions on the second derivatives of adjacent intervals must contradict one another. In turn, the involved first derivatives are decreased towards zero by a predetermined step size.}}}
    \item \indented{\indented{make\_monotone$(x_j,\ x_{j+1},\ f)$}}
    \item \indented{\indented{changed$(j) :=$ TRUE}}
      \indented{\indented{changed$(j+1) :=$ TRUE}} \label{step:changed}
    \item \indented{\indented{if $\big(j > 1$ and $(x_{j-1}, x_j) \not\in$ queue$\big)$ enqueue $\bigl((x_{j-1}, x_j)\bigr)$}}
      \indented{\indented{enqueue $\bigl((x_j, x_{j+1})\bigr)$}}
      \indented{\indented{if $\big(j+1 < n$ and $(x_{j+1}, x_{j+2}) \not\in$ queue$\big)$ enqueue $\bigl((x_{j+1}, x_{j+2})\bigr)$}} \label{step:requeue}
      \indented{\indented{\comment{Step \ref{step:changed} records the endpoints of the current interval as having been changed, while \ref{step:requeue} adds adjacent intervals to \texttt{queue} that may have inadvertently been made nonmonotone by the changes to the present interval.}}}      
      \indented{end if}
      end while
}
%% ----------------------------------------------------------------------

@@

Given the potential for successive monotonicity fixes across all intervals, the worst case runtime of Algorithm \ref{alg:monotone-spline} is $\mathcal{O}(s n)$ for $n$ breakpoints and shrink factor $s$. In practice this worst case has been observed to be unlikely.

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\textwidth]{cubic-quintic-sin}
  \caption{Depicted above are the monotone cubic and quintic spline interpolants of the function $\sin(x) + x$ over the points $[0, (5/2) \pi]$.  Notice that the maximum error of the cubic interpolant is larger, because it only captures first derivative information at the points. The quintic interpolant captures both first and second derivative information at the points.}
  \label{fig:cubic_quintic_sin}
\end{figure}

@@

\begin{table}
  \centering
  \begin{tabular}{c|c c c|c c c|c c c}
    \hline
    \multirow{2}{*}{$n$}
      & \multicolumn{3}{c|}{Checks} & \multicolumn{3}{c}{Fixes} & \multicolumn{3}{|c}{Search Steps} \\
      & min & median & max & min & median & max & min & median & max \\
    \hline
    5 & 4 & 6 & 274 & 0 & 1 & 154 & 3 & 31 & 2006\\
    25 & 28 & 47 & 378 & 2 & 14 & 225 & 80 & 348 & 3742\\
    50 & 61 & 124 & 484 & 6 & 40 & 238 & 217 & 851 & 4445\\
    75 & 102 & 216 & 842 & 14 & 74 & 411 & 463 & 1491 & 6333\\
    100 & 142 & 306 & 776 & 21 & 107 & 380 & 629 & 2140 & 7414\\
    \hline
  \end{tabular}
  \caption{Here we see a table of the minimum, median, and maximum number of checks, fixes, and binary search steps required in the execution of \texttt{make\_spline} for increasing size sequences, $n$, over $100$ randomly generated sets of monotone data. Notice the maximum for each counter is often significantly greater than the minimum and median, because the distribution of each counter is skewed right.}
  \label{table:e3_results}
\end{table}

@@

The monotone quintic spline interpolant provides a distinct increase in accuracy over the monotone cubic variant. Although the complexity of the algorithm is increased, the number of points required to achieve the same level of accuracy has decreased. In the present state, the algorithm for enforcing monotonicity on a spline is not as trivially parallel as the cubic algorithm, however it can still be parallelized.

Acknowledging that the algorithm for constructing a monotone quintic spline interpolant is slightly more complex than the cubic case, gains in accuracy or a decreased necessary number of points are often worth the computational effort.

@@

This paper proposes and tests an algorithm for constructing monotone quintic spline interpolants. Experiments demonstrate an improvement in approximation accuracy over monotone cubic spline interpolants, as expected based on theory.

There are still open avenues of research going forward, such as an alternative sufficient condition for enforcing monotonicity or increased order monotone approximations. If the monotonicity conditions can be generalized to any order or made linear, the search for a monotone interpolating spline could potentially be formulated as a convex optimization problem. Finally, this work could be used to improve cumulative distribution function (CDF) estimates as well as predictive models that use CDFs.

%% \section*{Acknowledgments}
%% The NSF is acknowledged for its support of this project.

@@

\textbf{\uppercase{THOMAS C. H. LUX}} is a Ph.D. candidate in computer
science at Virginia Polytechnic Institute and State University
(VPI\&SU) studying under Dr. Layne Watson. His research interests
include computational science, approximation theory, optimization, and
artificial intelligence. His email address is \email{tchlux@vt.edu}.

@@

\textbf{\uppercase{Yili Hong}} (Ph.D., Iowa State, 2009) has interests in
engineering statistics, statistical modeling, and data analysis.

>>>>

%% ------ Package for adding bars into margin next to comments. ------
%% Command for changing the color of text within a block.
\usepackage{xcolor}   % Color package for coloring text.
\usepackage[normalem]{ulem} % For strikethrough text and the \sout{} command.
\usepackage[xcolor,rightbars]{changebar}
\renewcommand{\changebarsep}{5mm}   % Push the bar farther into the margin.
\renewcommand{\changebarwidth}{1mm} % Increase the width of the bar.
\cbcolor{red} % Set the color of the change bars to red.
%% \newcommand\changed[2][]{#2} % <- hidden
\newcommand\changed[2][]{{\cbstart{\color{red}#2}\cbend}} % <- no old text
% \newcommand\changed[2][]{{\cbstart{\color{red}#2}{\color{gray}\sout{#1}}\cbend}}
%% -------------------------------------------------------------------

@@

A novel algorithm for computing monotone order six piecewise
polynomial interpolants is proposed. Algebraic constraints for
enforcing monotonicity are provided that align with quintic
monotonicity theory. The algorithm is implemented, tested, and applied
to several sample problems to demonstrate the improved accuracy of
monotone quintic spline interpolants compared to the previous
state-of-the-art monotone cubic spline interpolants.

%% An algorithm for computing monotone order six piecewise polynomial
%% interpolants is proposed. Algebraic constraints for enforcing
%% monotonicity are provided that align with quintic monotonicity
%% theory. \changed{The prior state of the art algorithm for constructing
%%   monotone spline interpolants is piecewise cubic, matching function
%%   values as well as first derivatives. The algorithm proposed here
%%   matches function value, first, and second derivatives achieving
%%   higher accuracy for $C^2$ functions while producing an approximation
%%   with continuous first derivative.} The \changed{proposed} algorithm
%% \changed{for constructing a monotone quintic interpolating spline} is
%% implemented, tested, and applied to several sample problems to
%% demonstrate the improved accuracy of monotone quintic spline
%% interpolants compared to that \changed{of} existing monotone cubic
%% spline interpolants.

@@

Many domains of science rely on smooth approximations to real-valued functions over a closed interval. Piecewise polynomial functions (splines) provide the smooth approximations for animation in graphics \cite{herman2015techniques,quint2003scalable}, aesthetic structural support in architecture \cite{brennan2019measure}, efficient aerodynamic surfaces in automotive and aerospace engineering \cite{brennan2019measure}, prolonged effective operation of electric motors \shortcite{berglund2009planning}, and accurate nonparametric approximations in statistics \cite{knott2012interpolating}. While polynomial interpolants and regressors apply broadly, splines are often a good choice because they can approximate globally complex functions while minimizing the local complexity of an approximation.

@@

The importance of piecewise quintic interpolation over lower order approximations can be simply demonstrated. In general, the order of a polynomial determines the number of function values it can interpolate, and the growth rate of error away from the interpolated function values. As demonstrated in Figure \ref{fig:spline_demonstration}, it can be seen that matching a value at either end of the interval requires an order two (linear) approximation and each additional derivative at the ends of the interval raises the necessary polynomial order by two. Hence, only even order (odd degree) interpolating splines are broadly applicable. The body of this work is composed of a novel algorithm for enforcing monotonicity on quintic polynomial pieces, then extending that solution to work on quintic splines.

@@

The current state-of-the-art monotone interpolating spline with a mathematical software implementation is piecewise cubic, continuously differentiable, and was first proposed in \cite{fritsch1980monotone} then expanded upon in \cite{carlson1985monotone}. Let $\pi: x_0 = k_1 < k_2 < \cdots < k_n = x_1$ be a partition of the interval $[x_0,x_1]$. Let $f: \mathbb{R} \rightarrow \mathbb{R}$ be $C^2$, and $\bigl\{f(k_i)\bigr\}_{i=1}^n$ and $\bigl\{\Delta_i\bigr\}_{i=1}^n$ be given sets of function and derivative values at the partition points for a monotone function $f$. Either $f(k_i) \leq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\ge0$, $i=1$, $\ldots$, $n$, or $f(k_i) \geq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\le0$, $i=1$, $\ldots$, $n$. Let $\hat f$ be a piecewise cubic polynomial defined in each subinterval $I_i = [k_i, k_{i+1}]$ by
\begin{align*}
h_i &= k_{i+1} - k_{i}, \quad
u(t) = 3t^2 - 2t^3, \quad
p(t) = t^3 - t^2,\\
\hat f(x) &= f(k_i) u\big((k_{i+1} - x) / h_i\big) +
f(k_{i+1}) u\big((x - k_i) / h_i\big) \\
&{} - h_i\Delta_i p\big((k_{i+1}-x)/h_i\big) + h_i\Delta_{i+1}
p\big((x-k_i)/h_i\big).
\end{align*}
Notice that a trivially monotone spline results when $\Delta_i = 0$, for $i = 1$, $\ldots$, $n$. However, such a spline has too many \textit{wiggles} for most applications. \citeN{carlson1985monotone} show that simple conditions on the derivative values can guarantee monotonicity, and that these conditions can be enforced in a way that ensures modifications on one interval will not break the monotonicity of cubic polynomials over any neighboring intervals. If $f(k_i) = f(k_{i+1})$, take $\Delta_i = \Delta_{i+1} =0$ and $\alpha =\beta =1$, otherwise let $\alpha = \big(\Delta_i (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$ and $\beta = \big(\Delta_{i+1}(k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$. Monotonicity of a cubic polynomial over a subinterval can be maintained by ensuring that $\alpha$ and $\beta$ reside in any of the regions depicted in Figure \ref{fig:projection}.

\begin{figure}
  \centering
  \includegraphics[width=0.40\textwidth]{cubic_projection_demonstration}
  \caption{These are the feasible regions of monotonicity for cubic splines and the projections that make a cubic polynomial piece monotone. The regions themselves are numbered 1--4 corresponding to their original description in Fritsch and Carlson (1980). One point is projected onto each region as a demonstration.}\label{fig:projection}
\end{figure}

@@

The following section is composed of three algorithms that together are used to construct a monotone quintic spline interpolant. Without loss of generality, the algorithms will only consider the monotone increasing (nondecreasing) case. The monotone decreasing case is handled similarly. Algorithm \ref{alg:check-monotone} checks monotonicity,
Algorithm \ref{alg:make-monotone} enforces monotonicity on an order six
polynomial piece of the quintic spline, and Algorithm
\ref{alg:monotone-spline} uses the previous two algorithms to enforce
monotonicity for the entire quintic spline.

@@

These algorithms make use of first and second derivatives of the approximated function where $D$ denotes the differentiation operator. In the case that the first and second derivative information is not provided along with function values, the derivatives are estimated with finite differences of the function values. The final quintic spline is represented as a piecewise polynomial using the Newton form for each polynomial piece, and ultimately converted to a $B$-spline representation for evaluation.

@@

Notice that both Algorithms \ref{alg:check-monotone} and \ref{alg:make-monotone} have $\mathcal{O}(1)$ runtime assuming a fixed level of precision is chosen beforehand. A constant number of operations are needed for verifying monotonicity (Algorithm \ref{alg:check-monotone}), while a constant set of operations and a single binary search are performed for enforcing monotonicity (Algorithm \ref{alg:make-monotone}). The search requires a fixed number of steps to achieve any predetermined relative precision, since its accuracy is predetermined and not a function of the problem at hand. Also note that an efficient implementation of Algorithm \ref{alg:make-monotone} only needs to recalculate Steps \ref{alg1-abg} and \ref{alg1-last} of Algorithm \ref{alg:check-monotone} during the binary search. Next, an algorithm for constructing a monotone quintic spline interpolant is presented.

@@

\algorithm
    {monotone\_spline$\big((k_1$, $\ldots$, $k_n)$, $f$, $s\big)$}
    {where $(k_1$, $\ldots$, $k_n)$ is an increasing sequence of real numbers, $f$ is an order six piecewise polynomial with breakpoints $k_1$, $\ldots$, $k_n$ defined by the data $\bigl\{f(k_i)\bigr\}_{i=1}^n$, $\bigl\{Df(k_i)\bigr\}_{i=1}^n$, $\bigl\{D^2f(k_i)\bigr\}_{i=1}^n$, and $s > 1$ is an integer shrink factor.}
    {\label{alg:monotone-spline}
    \item create queue
      \\ create changed$(1$, $\ldots$, $n) := $ FALSE
      \\ create step\_size$(1$, $\ldots$, $n) := \bigl(Df(k_1)/s$, $\ldots$, $Df(k_n)/s\bigr)$ \label{step:create}
      \comment{The three variables defined in Step \ref{step:create} are used to ensure the eventual achievement of monotonicity. The \texttt{queue} is a standard first in first out (FIFO) queue with \texttt{enqueue} and \texttt{dequeue} operations. The array \texttt{changed} contains Booleans describing whether or not a breakpoint belongs to an interval that has been modified to enforce monotonicity. The \texttt{step\_size} array contains the real-valued derivative decrement step sizes to use in the search for a valid monotone spline.}
    \item for $i := 1$, $\ldots$, $n-1$
      \indented{enqueue $\bigl((k_i, k_{i+1})\bigr)$}
      end for
      \comment{Initially, all intervals must be checked for monotonicity. The following loop will continue until all intervals are verified as monotone without need for modification.}
    \item while $($size$($queue$) > 0)$
    \item \indented{$(k_j$, $k_{j+1}) := $ dequeue}
    \item \indented{if $\big($not is\_monotone$(k_j$, $k_{j+1}$, $f)\big)$}
    \item \indented{\indented{if $\big($changed$(j)\big) \qquad \ Df(k_j) := Df(k_j) -{}$step\_size$(j)$}}
      \indented{\indented{if $\big($changed$(j+1)\big) \quad Df(k_{j+1}) := Df(k_{j+1}) -{}$step\_size$(j+1)$}}
      \indented{\indented{\comment{If a breakpoint belongs to an interval that has been previously modified, then the enforced monotonicity conditions on the second derivatives of adjacent intervals must contradict one another. In turn, the involved first derivatives are decreased towards zero by a predetermined step size.}}}
    \item \indented{\indented{make\_monotone$(k_j,\ k_{j+1},\ f)$}}
    \item \indented{\indented{changed$(j) :=$ TRUE}}
      \indented{\indented{changed$(j+1) :=$ TRUE}} \label{step:changed}
    \item \indented{\indented{if $\big(j > 1$ and $(k_{j-1}, k_j) \not\in$ queue$\big)$ enqueue $\bigl((k_{j-1}, k_j)\bigr)$}}
      \indented{\indented{enqueue $\bigl((k_j, k_{j+1})\bigr)$}}
      \indented{\indented{if $\big(j+1 < n$ and $(k_{j+1}, k_{j+2}) \not\in$ queue$\big)$ enqueue $\bigl((k_{j+1}, k_{j+2})\bigr)$}} \label{step:requeue}
      \indented{\indented{\comment{Step \ref{step:changed} records the endpoints of the current interval as having been changed, while \ref{step:requeue} adds adjacent intervals to \texttt{queue} that may have inadvertently been made nonmonotone by the changes to the present interval.}}}      
      \indented{end if}
      end while
}
%% ----------------------------------------------------------------------

@@

If successive monotonicity fixes are required for all intervals, the worst case runtime of Algorithm \ref{alg:monotone-spline} is $\mathcal{O}(s n)$ for $n$ breakpoints and shrink factor $s$. In practice this worst case has been observed to be unlikely.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.7\textwidth]{cubic-quintic-sin}
  \caption{Depicted above are the monotone cubic and quintic spline interpolants of the function $\sin(x) + x$ over the points $[0, (5/2) \pi]$.  Notice that the maximum error of the cubic interpolant is larger, because it only captures first derivative information at the points. The quintic interpolant captures both first and second derivative information at the points.}
  \label{fig:cubic_quintic_sin}
\end{figure}

@@

\begin{table}
  \centering
  \begin{tabular}{c|c c c|c c c|c c c}
    \hline
    \multirow{2}{*}{$n$}
      & \multicolumn{3}{c|}{Checks} & \multicolumn{3}{c}{Fixes} & \multicolumn{3}{|c}{Search Steps} \\
      & min & median & max & min & median & max & min & median & max \\
    \hline
    5 & 4 & 6 & 274 & 0 & 1 & 154 & 3 & 31 & 2006\\
    25 & 28 & 47 & 378 & 2 & 14 & 225 & 80 & 348 & 3742\\
    50 & 61 & 124 & 484 & 6 & 40 & 238 & 217 & 851 & 4445\\
    75 & 102 & 216 & 842 & 14 & 74 & 411 & 463 & 1491 & 6333\\
    100 & 142 & 306 & 776 & 21 & 107 & 380 & 629 & 2140 & 7414\\
    \hline
  \end{tabular}
  \caption{The minimum, median, and maximum number of checks, fixes, and binary search steps required in the execution of \texttt{make\_spline} for increasing size sequences, $n$, over $100$ randomly generated sets of monotone data. Notice the maximum for each counter is often significantly greater than the minimum and median, because the distribution of each counter is skewed right.}
  \label{table:e3_results}
\end{table}

@@

The monotone quintic spline interpolant provides a distinct increase in accuracy over the monotone cubic variant. The computational complexity of this algorithm is identical to the existing cubic method and the cost of evaluating the spline after construction is unchanged. While the number of computations required per interval has increased, the number of points required to achieve the same level of accuracy has decreased. In the present state, the algorithm for enforcing monotonicity on a spline is not as trivially parallel as the cubic algorithm, however it can still be parallelized. The checks for monotonicity across all intervals are independent and the monotonicity adjustments could be computed independently and intelligently merged with minor changes to the serial algorithm.

Acknowledging that the algorithm for constructing a monotone quintic spline interpolant is slightly more expensive than the cubic case, gains in accuracy or decreased necessary number of points are often worth the computational effort. In practice the costs of spline interpolation are often dominated by evaluation rather than construction, and the increased accuracy is afforded for a negligible increase in computational cost.

@@

This paper proposes and tests an algorithm for constructing monotone quintic spline interpolants. Experiments demonstrate an improvement in approximation accuracy over monotone cubic spline interpolants, as expected based on theory. There are still open avenues of research going forward, such as an alternative sufficient condition for enforcing monotonicity or increased order monotone approximations. If the monotonicity conditions can be generalized to any order or made linear, the search for a monotone interpolating spline could potentially be formulated as a convex optimization problem. Finally, this work could be used to improve cumulative distribution function (CDF) estimates as well as predictive models that use CDFs.

\section*{Acknowledgments}
This work was supported by the National Science Foundation Grants CNS-1565314 and CNS-1838271.

@@

\textbf{\uppercase{THOMAS C. H. LUX}} is a Ph.D. candidate in computer
science at Virginia Polytechnic Institute and State University
(VPI\&SU) studying under Dr. Layne Watson. His email address is
\email{tchlux@vt.edu}.

@@

\textbf{\uppercase{Yili Hong}} (Ph.D., Iowa State, 2009) is an
associate professor of Statistics at VPI\&SU.

### 2020-03-30 [2020-01]_IEEESE_(convex_optimization)/paper.tex
updated work

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%2345678901234567890123456789012345678901234567890123456789012345678901234567890
%        1         2         3         4         5         6         7         8

\documentclass[conference]{ieeeconf}  % Comment this line out
                                                          % if you need a4paper
%\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4 paper
\IEEEoverridecommandlockouts                              % This command is only needed 
%% % copyright info
%% \IEEEpubid{\makebox[\columnwidth]{978-1-5386-6133-8/18/\$31.00
%%         ~\copyright~2018 IEEE \hfill}\hspace{\columnsep}
%%         \makebox[\columnwidth]{ }}

% if you want to use the \thanks command
\overrideIEEEmargins
% See the \addtolength command later in the file to balance the column lengths on the last page of the document

% The following packages can be found on http:\\www.ctan.org
\usepackage{tikz}     % For adding axes to plots
\usepackage{graphicx} % For including PDFs
\usetikzlibrary{positioning}
\usepackage{balance} % For balancing the last column.

%% \usepackage{graphics} % for pdf, bitmapped graphics files
%\usepackage{epsfig} % for postscript graphics files
%\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
\usepackage{amsmath} % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed

@@

@@

%% \author{Thomas C. H. Lux$^{1}$, Tyler H. Chang$^{1}$% <-this % stops a space
%% \thanks{$^{1}$Doctoral candidate, Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu, thchang at vt.edu)}}%
%% }
\author{Anonymous Authors}

@@

\maketitle
\thispagestyle{empty}
\pagestyle{empty}

>>>>

\documentclass[letterpaper, 10 pt, conference]{IEEEtran}  % Comment this line out
                                                          % if you need a4paper

% The following packages can be found on http:\\www.ctan.org
\usepackage{tikz}     % For adding axes to plots
\usepackage{graphicx} % For including PDFs
\usetikzlibrary{positioning}
\usepackage{balance} % For balancing the last column.
\usepackage{amsmath} % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed

@@

%% ------------------------------------------
%% Block to make a ``titleheader''
\makeatletter
\newcommand*\titleheader[1]{\gdef\@titleheader{#1}}
\AtBeginDocument{%
  \let\st@red@title\@title%
  \def\@title{%
    \bgroup\normalfont\large\raggedleft\@titleheader\par\egroup
    \vskip1.5em\st@red@title}
}
\makeatother
%% ------------------------------------------

@@

\titleheader{IEEE SoutheastCon 2020}

\author{Thomas C. H. Lux$^{1}$, Tyler H. Chang$^{1}$% <-this % stops a space
\thanks{$^{1}$Doctoral candidate, Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu, thchang at vt.edu)}}%
}

@@

%\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4 paper
\IEEEoverridecommandlockouts                              % This command is only needed 
%% % copyright info
\IEEEpubid{\makebox[\columnwidth]{978-1-7281-6861-6/20/\$31.00~\copyright{}2020 IEEE \hfill} \hspace{\columnsep}\makebox[\columnwidth]{ }}

\maketitle
